Developmental integration cannot explain major features of stomatal anatomical evolution in seed plants

### 2021-09-28 ms/ms.tex
initial amnat submission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  10pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{geometry}
\geometry{verbose,letterpaper,top=0.85in,left=2.75in,footskip=0.75in,marginparwidth=2in}

% \usepackage[breaklinks=true,pdfstartview=FitH,citecolor=blue]{hyperref}
\hypersetup{colorlinks,%
	citecolor=blue,%
	filecolor=red,%
	linkcolor=blue,%
	urlcolor=red,%
	pdfstartview=FitH}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{textgreek}
\usepackage[greek,english]{babel}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage[osf]{libertine}
\usepackage{libertinust1math}
\usepackage{inconsolata}

\usepackage{booktabs}

\usepackage{setspace}
\doublespacing

% \setstretch{1.8999999999999999}

\usepackage[right]{lineno}

% \renewcommand{\rmdefault}{cmr}

% Additional template based on: https://www.overleaf.com/project/5dba26749d2ef50001d3a002

% clean citations
\usepackage{cite}

% hyperref makes references clicky. use \url{www.example.com} or \href{www.example.com}{description} to add a clicky url
\usepackage{nameref,hyperref}

% improves typesetting in LaTeX
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% text layout - change as needed
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% adjust caption style
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}

% remove brackets from references
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% headrule, footrule and page numbers
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}

% use \textcolor{color}{text} for colored text (e.g. highlight to-do areas)
\usepackage{color}

% define custom colors (this one is for figure captions)
\definecolor{Gray}{gray}{.25}

% this is required to include graphics
\usepackage{graphicx}

% use if you want to put caption to the side of the figure - see example in text
\usepackage{sidecap}

% use for have text wrap around figures
\usepackage{wrapfig}
\usepackage[pscoord]{eso-pic}
\usepackage[fulladjust]{marginnote}
\reversemarginpar

% flush left while keep identation
\makeatletter
\newcommand\iraggedright{%
  \let\\\@centercr\@rightskip\@flushglue \rightskip\@rightskip
  \leftskip\z@skip}
\makeatother

% make pdf as default figure format
\DeclareGraphicsExtensions{.pdf,.png, %
    .jpg,.mps,.jpeg,.jbig2,.jb2,.JPG,.JPEG,.JBIG2,.JB2}

\usepackage{newunicodechar}
\DeclareRobustCommand{\okina}{%
  \raisebox{\dimexpr\fontcharht\font`A-\height}{%
    \scalebox{0.8}{`}%
  }%
}
\newunicodechar{ʻ}{\okina}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\author{}
\date{\vspace{-2.5em}}

@@

% align only at left, not at right.
\iraggedright

\setstretch{1.5}
\begin{flushleft}
{\Large
\textbf\newline{Developmental integration cannot explain major features of stomatal anatomical evolution in seed plants}
}
\newline
\\
Christopher D. Muir\textsuperscript{1,*},
Miquel \`{A}ngel Conesa\textsuperscript{2},
Jeroni Galm\'{e}s\textsuperscript{2},
Varsha S. Pathare\textsuperscript{3},
Patricia Rivera\textsuperscript{4},
Rosana López Rodríguez\textsuperscript{5},
Teresa Terrazas\textsuperscript{4},
Dongliang Xiong\textsuperscript{6}
\\
\bigskip
\bf{1} School of Life Sciences, University of Hawaiʻi at M\=anoa, Honolulu, HI 96822, USA \\
\bf{2} Research Group on Plant Biology under Mediterranean Conditions, Departament de Biologia, Universitat de les Illes Balears, Ctra. Valldemossa km 7.5, E-07122, Palma, Spain \\
\bf{3} School of Biological Sciences, Washington State University, Pullman, WA 99164-4236, USA \\
\bf{4} Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Apartado Postal 70‑367, 04510 Mexico City, Mexico \\
\bf{5} Departamento de Sistemas y Recursos Naturales, Universidad Politécnica de Madrid, 28040 Madrid, Spain \\
\bf{6} National Key Laboratory of Crop Genetic Improvement, MOA Key Laboratory of Crop Ecophysiology and Farming System in the Middle Reaches of the Yangtze River, College of Plant Science and Technology, Huazhong Agricultural University, Wuhan, Hubei 430070, China\\
\bigskip
* cdmuir@hawaii.edu

\end{flushleft}

@@

\begin{itemize}
\tightlist
\item
  Developmental integration can cause traits to covary over macroevolutionary time and in some cases prevent populations from reaching their adaptive optima. Developmental integration between stomatal size and density may contribute to two major features of stomatal anatomical evolution: inverse size-density scaling and bimodal stomatal ratio. If these patterns result from developmental integration, we predicted that in amphistomatous leaves 1) stomatal size and density should covary similarly on both abaxial and adaxial surfaces and 2) stomatal traits (size and density) on each surface should covary isometrically.
\item
  We synthesized data on stomatal density and length from amphistomatous leaves of 711 terrestrial seed plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from 327 phylogenetically independent contrasts using a robust Bayesian model.
\item
  Adaxial stomatal density, but not length, is evolutionarily labile and not strongly integrated with stomatal length or abaxial stomatal density. Hence, developmental integration alone cannot explain inverse size-density scaling nor bimodal stomatal ratio.
\item
  Quasi-independent evolution of stomatal anatomical traits facilitates largely unfettered access to fitness optima. If stomatal anatomical traits are near their current fitness optimum, this implies that limits on trait (co)variance result from selective rather than developmental constraints. However, we cannot rule out that developmental integration is important in some lineages. Future research should identify the mechanistic basis of (dis)integration in stomatal development.
\end{itemize}

\textbf{Keywords:} Adaptation, amphistomy, developmental integration, leaf, phylogenetic comparative methods, stomata

\linenumbers

@@

We used the R packages \textbf{taxonlookup} version 1.1.5 (\protect\hyperlink{ref-pennell_simple_2016}{Pennell, FitzJohn, and Cornwell 2016}) and \textbf{V.phylomaker} version 0.1.0 (\protect\hyperlink{ref-jin_vphylomaker_2019}{Jin and Qian 2019}) to maximize overlap between our data set and the GBOTB.extended mega-tree of seed plants (\protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018} ; \protect\hyperlink{ref-zanne_three_2014}{Zanne et al. 2014}). We further resolved large (\(\ge4\) taxa) polytomies in 30 clades with sufficient sequence data using \textbf{PyPHLAWD} version 1.0 (\protect\hyperlink{ref-smith_pyphlawd_2019}{S. A. Smith and Walker 2019}) in Python 3.8.2 (Python Software Foundation, \url{https://www.python.org/}). We used sequence data from the most recent GenBank Plant and Fungal sequences database division (\protect\hyperlink{ref-ouellette_database_1997}{Ouellette and Boguski 1997}). We inferred subtree phylogenies using RAxML version 8.2.12 (\protect\hyperlink{ref-stamatakis_raxml_2014}{Stamatakis 2014}) and conducated molecular dating using the \texttt{chronos()} function in the R package \textbf{ape} version 5.5 (\protect\hyperlink{ref-paradis_ape_2019}{Paradis and Schliep 2019}) to obtain ultrametric trees. We grafted resolved, ultrametric subtrees onto the mega-tree at the polytomy nodes and rescaled to keep the mega-tree ultrametric. In some cases, resolving polytomies was not possible because there was little or no overlap between taxa in the data set and taxa with sequence data available for \textbf{PyPHLAWD}. In these cases, we randomly selected two taxa as a phylogenetially independent pair and dropped the rest. Remaining polytomies of three taxa were resolved randomly using the \texttt{multi2di()} function in \textbf{ape}. The final data set for which we had both trait and phylogenetic information contained 711 taxa (Notes S1). Seven taxa are gymnosperms; the vast majority are angiosperms.

@@

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

We thank Jacob Watts for comments on earlier versions of this manuscript.

\hypertarget{author-contribution}{%
\section{Author Contribution}\label{author-contribution}}

CDM designed the study, compiled data, analyzed data, and wrote the manuscript with input from all authors. All authors contributed data.

@@

The final data set and phylogeny used in the analysis are included in the \protect\hyperlink{supporting-information}{Supporting Information}. The raw anatomical data and source code will be archived on Zenodo upon publication.

@@

\hypertarget{supporting-information}{%
\section{Supporting Information}\label{supporting-information}}

\renewcommand\thefigure{S\arabic{figure}}    
\renewcommand\thetable{S\arabic{table}}    
\setcounter{figure}{0}    
\setcounter{table}{0}

\textbf{Note S1:} Final fully resolved bifurcating phylogenetic tree of 711 taxa in Newick format.

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% \documentclass[11pt]{article} % redundant
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

\title{Developmental integration cannot explain major features of stomatal anatomical evolution in seed plants}

\setlength{\parindent}{5ex}

\author{}
\date{\vspace{-2.5em}}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\author{}
\date{}

@@

\setstretch{1.7}
\maketitle

\bigskip

\textit{Manuscript elements}: Figure 1, figure 2, figure 3, figure 4, figure 5, table 1, online appendix A (including Notes A1, table A1, table A2, and table A3).

\bigskip

\textit{Keywords}: adaptation, amphistomy, developmental integration, leaf, phylogenetic comparative methods, stomata

\bigskip

\textit{Manuscript type}: Major Article

\bigskip

\textit{Word Count}: 4,704

\bigskip

\linenumbers{}
\modulolinenumbers[3]

\newpage{}

@@

Developmental integration can cause traits to covary over macroevolutionary time and in some cases prevent populations from reaching their adaptive optima. Developmental integration between stomatal size and density may contribute to two major features of stomatal anatomical evolution: inverse size-density scaling and bimodal stomatal ratio. If these patterns result from developmental integration, we predicted that in amphistomatous leaves 1) stomatal size and density should covary similarly on both abaxial and adaxial surfaces and 2) stomatal traits (size and density) on each surface should covary isometrically. We synthesized data on stomatal density and length from amphistomatous leaves of 711 terrestrial seed plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from 327 phylogenetically independent contrasts using a robust Bayesian model. Adaxial stomatal density, but not length, is evolutionarily labile and not strongly integrated with stomatal length or abaxial stomatal density. Hence, developmental integration alone cannot explain inverse size-density scaling nor bimodal stomatal ratio. Quasi-independent evolution of stomatal anatomical traits facilitates largely unfettered access to fitness optima. If stomatal anatomical traits are near their current fitness optimum, this implies that limits on trait (co)variance result from selective rather than developmental constraints. However, we cannot rule out that developmental integration is important in some lineages. Future research should identify the mechanistic basis of (dis)integration in stomatal development.

\newpage{}

@@

We used the R packages \textbf{taxonlookup} version 1.1.5 (\protect\hyperlink{ref-pennell_simple_2016}{Pennell, FitzJohn, and Cornwell 2016}) and \textbf{V.phylomaker} version 0.1.0 (\protect\hyperlink{ref-jin_vphylomaker_2019}{Jin and Qian 2019}) to maximize overlap between our data set and the GBOTB.extended mega-tree of seed plants (\protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018} ; \protect\hyperlink{ref-zanne_three_2014}{Zanne et al. 2014}). We further resolved large (\(\ge4\) taxa) polytomies in 30 clades with sufficient sequence data using \textbf{PyPHLAWD} version 1.0 (\protect\hyperlink{ref-smith_pyphlawd_2019}{S. A. Smith and Walker 2019}) in Python 3.8.9 (Python Software Foundation, \url{https://www.python.org/}). We used sequence data from the most recent GenBank Plant and Fungal sequences database division (\protect\hyperlink{ref-ouellette_database_1997}{Ouellette and Boguski 1997}). We inferred subtree phylogenies using RAxML version 8.2.12 (\protect\hyperlink{ref-stamatakis_raxml_2014}{Stamatakis 2014}) and conducated molecular dating using the \texttt{chronos()} function in the R package \textbf{ape} version 5.5 (\protect\hyperlink{ref-paradis_ape_2019}{Paradis and Schliep 2019}) to obtain ultrametric trees. We grafted resolved, ultrametric subtrees onto the mega-tree at the polytomy nodes and rescaled to keep the mega-tree ultrametric. In some cases, resolving polytomies was not possible because there was little or no overlap between taxa in the data set and taxa with sequence data available for \textbf{PyPHLAWD}. In these cases, we randomly selected two taxa as a phylogenetially independent pair and dropped the rest. Remaining polytomies of three taxa were resolved randomly using the \texttt{multi2di()} function in \textbf{ape}. The final data set for which we had both trait and phylogenetic information contained 711 taxa (Notes A1). Seven taxa are gymnosperms; the vast majority are angiosperms.

@@

@@

The final data set and phylogeny used in the analysis are included in the {[}Supporting Information{]}. The raw anatomical data and source code will be archived on Zenodo upon publication.

@@

\hypertarget{online-appendix}{%
\section{Online Appendix}\label{online-appendix}}

\renewcommand\thefigure{A\arabic{figure}}    
\renewcommand\thetable{A\arabic{table}}    
\setcounter{figure}{0}    
\setcounter{table}{0}

@@

\hypertarget{literature-cited}{%
\section*{Literature Cited}\label{literature-cited}}
\addcontentsline{toc}{section}{Literature Cited}
